A load balancing scheme for massively multiplayer online games \thanks{This work was supported by the National Research Council (CNPq) and by the Coordination of Improvement of Higher Education (CAPES), both brazilian research funding agencies.}

### 2009-04-10 trunk/textos/dissertation/MTAP_LaTeX/pub/tex/latex/svjour3/global/balancing.tex

\newcommand{\ggp}{ProGReGA}
%\newcommand{\ggp}{PGRGA}
\newcommand{\ggpmeaning}{proportional greedy region growing algorithm}
\newcommand{\ggpk}{\mbox{\ggp-KH}}
\newcommand{\ggpkmeaning}{proportional greedy region growing algorithm keeping heaviest cell}
\newcommand{\ggpf}{\ggp-KF}
\newcommand{\ggpfmeaning}{proportional greedy region growing algorithm keeping usage fraction}
\newcommand{\bfa}{BFBCT}
\newcommand{\bfameaning}{best-fit based cell transference}
\newcommand{\kl}{Kernighan\-Lin}
\newcommand{\klalg}{Kernighan-Lin}
\newcommand{\wtodiv}{weight\_to\_divide}
\newcommand{\freecap}{free\_capacity}
\newcommand{\rlist}{region\_list}
\newcommand{\clist}{cell\_list}
\newcommand{\locgroup}{local\_group}
\newcommand{\avguse}{average\_usage}
\newcommand{\wlocal}{local\_weight}
\newcommand{\caplocal}{local\_capacity}
%\newcommand{\wshare}{por\text{\emph{\c{c}}}\tilde{a}o\_da\_carga}
\newcommand{\wshare}{weight\_share}
\newcommand{\wlose}{weight\_to\_lose}
\newcommand{\destreg}{destination\_regions}
\newcommand{\regcap}{free\_capacity}
\newcommand{\wtoreg}{weight\_to\_this\_region}
\newcommand{\misccite}[2]{Available at: #2}
\newcommand{\gamecite}[2]{Available at: #2}

@@

It is considered that an initial division of the virtual environment has already been made. Each server should then check regularly if there is an imbalance and trigger the algorithm. While the overhead resulting from the distribution of virtual environment is part of the workload on servers, there is no way to know it beforehand without executing the repartitioning first. For this reason, the ``weight'' to be distributed does not include this extra overhead.

When there is an unbalanced region, it is selected a local group of regions, similar to what was shown in section \ref{sec:local}, but with some changes (section \ref{sec:alg:localselect}). After this selection, it will be used an algorithm whose parameters are only the loads of the cells and their interactions, because such data is available. Finally, with the regions already balanced, they will be mapped on a graph and the algorithm of Kernighan and Lin \cite{kernighan1970ehp} will be used to refine the partitioning, reducing the edge-cut and, thus, the overhead, while keeping the balance.

@@

The algorithms were developed oriented to regions, instead of servers, in order to be more legible, becaus of the constant transfers of cells. Moreover, it becomes easier to extend, in the future, the balancing model used here to allow more than one server managing the same region. The proposed algorithms are described as follows. The one in section \ref{sec:alg:localselect} is for phase 1, the sections \ref{sec:alg:ggp} to \ref{sec:alg:bfa} are alternatives for phase 2 and the algorithm in section \ref{sec:alg:kl} is the refinement of phase 3.

@@

A possible undesirable effect of the \ggp\ algorithm is that by releasing all the cells and redistributing them, it might happen that one or more regions completely change their place, causing several players to disconnect from their servers and reconnect to a new one. To try to reduce the likelihood of such event, we propose a variation of the algoritgh, called, \textbf{\ggpk} (\ggpkmeaning). This new algorithm (shown in Algorithm \ref{alg:ggpk}) is similar to the original version, except that each region maintains its heaviest cell (lines \ref{alg:ggpk:saveheaviestcell} and \ref{alg:ggpk:getbackheaviestcell}), from which a region similar to the previous one can be formed, so that several players will not need to migrate to other server. However, to keep one of the cells of each region, it might be preventing a better balancing to occur. Also, the fixation of that cell might cause fragmented regions, increasing the overhead.

@@

The \textbf{\bfa} (\bfameaning) is proposed here as an alternative to \ggp\ and its variants. The objective of the algorithm is to check what is the weight excess in each region and transfer it to free regions whose capacity is the closest to that value. This is done by transferring cells whose weight is the closest to free capacity of the receiving region, observed two restrictions: first, the total weight transferred can not be larger than the free capacity of the destination region and, second, we should not transfer a load greater than the necessary to eliminate the overload. The second restriction is justified because a weight transfer larger than necessary would probably result in a larger amount of migrating players. The exception to this rule is when a cell is heavier than the other, not for having more avatars, but because they are closer, with quadratic traffic growth between them. The Algorithm \ref{alg:bfa} describes in detail the operation of \bfa.

@@

After balancing the load between servers in phase 2, each server will have a usage ratio similar to the others. However, phase 2 does not consider the interaction between cells and the overhead between regions, which may have been fragmented, resulting in many pairs of neighboring cells that are not part of the same region. This increases the probability that two players, each in one of these cells, interact with each other close to the border, which, as was shown in previous sections, causes a waste of resources of the server system.

To minimize this problem, it is proposed here to use the algorithm of Kernighan and Lin \cite{kernighan1970ehp}. Although the algorithm receives as input a graph, the vertices, edges, weights and partitions can be interpreted respectively as cells, interactions, loads and regions. Given a pair of regions, \kl\ searches for pairs of cells that, when exchanged between their regions, the overhead is reduced.

Let $R_A$ $R_B$ be two regions and $a$ a cell such that $a \in R_A$. The external interaction of $a$ is defined as $E(a) = \sum_{b \in R_b} Int_c (a, b)$ and the internal interaction is defined as $I(a) = \sum_{a' \in R_A} Int_c(a, a')$. The cell $a$ is prone to change regions if its interaction with $R_B$ is greater than its interaction with $R_A$, i.e. $D(a) = E(a) - I(a) > 0$. Assuming that there is a cell $b \in R_B$ such that $D(a) + D(b) - 2 \times Int_c(a, b) > 0 $, the exchange will reduce the overhead between the regions. The term $D(a) + D(b) - 2 \times Int_c(a, b)$ is refered as $gain(a, b)$, because it represents the gain (reduction of overhead) of exchanging the regions of $a$ and $b$.

In the case of a virtual environment distributed among various regions, generally more than two, \kl\ is run for each pair of regions. Moreover, after each exchange the balance should be ketp -- otherwise, all cells could go to the same region, completely eliminating the overhead. The \kl\ is widely known in the area of distributed systems, so detail will not be provided here. Consider only that it returns a value of true if any change was made and false if it not. It runs for all pairs of regions. until it returns a value of false, indicating that no exchange will provide additional gain.
The Algorithm \ref{alg:kl} shows how the algorithm \kl\ is called.

@@

	\STATE $swapped \leftarrow$ \textbf{true}
	\WHILE{$swapped =$ \textbf{true}}
		\STATE $swapped \leftarrow$ \textbf{false}
		\FOR{each region $R_i$ in $\rlist$}
			\FOR{each region $R_j$ in $\rlist$}
				\IF{$\kl(R_i, R_j) =$ \textbf{true}}
					\STATE $swapped \leftarrow$ \textbf{true}
				\ENDIF
			\ENDFOR
		\ENDFOR
	\ENDWHILE

>>>>

\newcommand{\ggp}{ProGReGA}
%\newcommand{\ggp}{PGRGA}
\newcommand{\ggpmeaning}{proportional greedy region growing algorithm}
\newcommand{\ggpk}{\mbox{\ggp-KH}}
\newcommand{\ggpkmeaning}{proportional greedy region growing algorithm keeping heaviest cell}
\newcommand{\ggpf}{\ggp-KF}
\newcommand{\ggpfmeaning}{proportional greedy region growing algorithm keeping usage fraction}
\newcommand{\bfa}{BFBCT}
\newcommand{\bfameaning}{best-fit based cell transference}
\newcommand{\kl}{Kernighan-Lin}
\newcommand{\klalg}{Kernighan-Lin}
\newcommand{\wtodiv}{weight\_to\_divide}
\newcommand{\freecap}{free\_capacity}
\newcommand{\rlist}{region\_list}
\newcommand{\clist}{cell\_list}
\newcommand{\locgroup}{local\_group}
\newcommand{\avguse}{average\_usage}
\newcommand{\wlocal}{local\_weight}
\newcommand{\caplocal}{local\_capacity}
%\newcommand{\wshare}{por\text{\emph{\c{c}}}\tilde{a}o\_da\_carga}
\newcommand{\wshare}{weight\_share}
\newcommand{\wlose}{weight\_to\_lose}
\newcommand{\destreg}{destination\_regions}
\newcommand{\regcap}{free\_capacity}
\newcommand{\wtoreg}{weight\_to\_this\_region}
\newcommand{\misccite}[2]{Available at: #2}
\newcommand{\gamecite}[2]{Available at: #2}

@@

It is considered that an initial division of the virtual environment has already been made. Each server should then check regularly if there is an imbalance and trigger the algorithm. Although the overhead resulting from the distribution of the virtual environment is part of the workload on servers, there is no way to know it beforehand without executing the repartitioning first. For this reason, the ``weight'' to be distributed does not include this extra overhead.

When there is an unbalanced region, it is selected a local group of regions, similar to what was shown in section \ref{sec:local}, but with some changes (section \ref{sec:alg:localselect}). After this selection, it will be used an algorithm whose parameters are only the loads of the cells and their interactions, for such data is available prior to distribute. Finally, with the regions already balanced, the algorithm of Kernighan and Lin \cite{kernighan1970ehp} will be used to refine the partitioning, reducing the edge-cut and, thus, the overhead, while keeping the balance.

@@

The algorithms were developed oriented to regions, instead of servers, in order to be more legible, because of the constant transfers of cells. Moreover, it becomes easier to extend, in the future, the balancing model used here to allow more than one server managing the same region. The proposed algorithms are described as follows. The one in section \ref{sec:alg:localselect} is for phase 1, the sections \ref{sec:alg:ggp} to \ref{sec:alg:bfa} are alternatives for phase 2 and the algorithm in section \ref{sec:alg:kl} is the refinement of phase 3.

@@

A possible undesirable effect of the \ggp\ algorithm is that by releasing all the cells and redistributing them, it might happen that one or more regions completely change their place, causing several players to disconnect from their servers and reconnect to a new one. To try to reduce the likelihood of such event, we propose a variation of the algorithm, called \textbf{\ggpk} (\ggpkmeaning). This new algorithm (shown in Algorithm \ref{alg:ggpk}) is similar to the original version, except that each region maintains its heaviest cell (lines \ref{alg:ggpk:saveheaviestcell} and \ref{alg:ggpk:getbackheaviestcell}), from which a region similar to the previous one can be formed, so that several players will not need to migrate to other server. However, to keep one of the cells of each region, it might be preventing a better balancing to occur. Also, the fixation of that cell might cause fragmented regions, increasing the overhead.

@@

The \textbf{\bfa} (\bfameaning) is proposed here as an alternative to \ggp\ and its variants. The objective of the algorithm is to check what is the weight excess in each region and transfer it to free regions whose capacity is the closest to that value. This is done by transferring cells whose weight is the closest to free capacity of the receiving region, observed two restrictions: first, the total weight transferred can not be larger than the free capacity of the destination region and, second, we should not transfer a load greater than the necessary to eliminate the overload. The second restriction is justified because a weight transfer larger than necessary would probably result in a larger amount of migrating players. The exception to this rule is when a cell is heavier than the other, not for having more avatars, but because they are closer, with quadratic traffic growth between them. Algorithm \ref{alg:bfa} describes in detail the operation of \bfa.

@@

After balancing the load between servers in phase 2, all the servers will have a similar resource usage ratio. However, the algorithm in phase 2 cannot measure the final interaction between regions before repartitioning. This may lead to a partitioning that, although every server uses a similar percentage of its upload bandwidth to its clients, there may be a high inter-server communication overhead, causing a waste of resources of the server system.

To attenuate this problem, it is proposed here to use the algorithm of Kernighan and Lin \cite{kernighan1970ehp}. This algorithm receives as input the graph that represents the virtual environment. Given a pair of regions (partitions), \kl\ searches for pairs of cells (vertices) that, when exchanged between their regions, the overhead (edge-cut) is reduced.

Let $R_A$ and $R_B$ be two regions and $a$ a cell such that $a \in R_A$. The external interaction of $a$ is defined as $E(a) = \sum_{b \in R_b} Int_c (a, b)$ and the internal interaction is defined as $I(a) = \sum_{a' \in R_A} Int_c(a, a')$. The cell $a$ is prone to change regions if its interaction with $R_B$ is greater than its interaction with $R_A$, i.e. $D(a) = E(a) - I(a) > 0$. Assuming that there is a cell $b \in R_B$ such that $D(a) + D(b) - 2 \times Int_c(a, b) > 0 $, the exchange will reduce the overhead between the regions. The term $D(a) + D(b) - 2 \times Int_c(a, b)$ is referred as $gain(a, b)$, because it represents the gain (reduction of overhead) of swapping the regions of $a$ and $b$.

In the case of a virtual environment distributed among various regions, generally more than two, \kl\ is run for each pair of regions. Moreover, after each swap performed by the algorithm, the balance must be kept -- otherwise, a completely unbalanced partitioning, with the lowest possible overhead, could be reached. The \kl\ algorithm is widely known in the area of distributed systems, so details will not be provided here. Consider only that it returns a value of $true$ if any change was made and $false$ otherwise. It runs for all pairs of regions. until it returns a value of false, indicating that no exchange will provide additional gain. Algorithm \ref{alg:kl} shows how the \kl\ algorithm is called.

@@

	\STATE $swapped \leftarrow$ \textbf{true}
	\WHILE{$swapped =$ \textbf{true}}
		\STATE $swapped \leftarrow$ \textbf{false}
		\FOR{each region $R_i$ in $\rlist$}
			\FOR{each region $R_j$ in $\rlist$}
				\IF{$Kernighan$-$Lin(R_i, R_j) =$ \textbf{true}}
					\STATE $swapped \leftarrow$ \textbf{true}
				\ENDIF
			\ENDFOR
		\ENDFOR
	\ENDWHILE
